---
title: "ANOVA"
geometry: margin = 2.5cm
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{lmodern}
- \usepackage{fancyhdr}
- \usepackage{booktabs}
- \usepackage{breqn}
- \pagestyle{fancy}
- \fancyhead{}
- \lhead{Report 1}
- \rhead{An, Li, Shi, Yu}
- \headheight = 15pt
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
  html_document:
    df_print: paged
pagenumbering: yes
fontsize: 12pt
toc: no
---


```{r setup, include=FALSE}
rm(list = ls(all.names = TRUE))
library(knitr)

#library(plotly)
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```


\section{Introduction}

sdfsdfsdfsdfs



\section{Exploratory Data Analysis}

```{r load data, include=FALSE}
potato<-read.table("./potato.dat.txt",col.names = c("area","temp","size","storage","cook", "texture", "flavour", "moist"))
# leave out texture and moist
summary(potato) # it is a balanced set
```

```{r EDA_plot}
layout(mat = matrix(c(1,1,2,2,3,3,
                      0,4,4,5,5,0), nrow = 2, byrow = TRUE))
boxplot(potato$flavour~potato$area, data = potato, xlab = "Growing Area", ylab = "Flavor score")
boxplot(potato$flavour~potato$temp, data = potato, xlab = "Temperature", ylab = "Flavor score")
boxplot(potato$flavour~ potato$size, data = potato, xlab = "Size of Potato", ylab = "Flavor score")
boxplot(potato$flavour~ potato$storage, data = potato, xlab = "Storage method", ylab = "Flavor score")
boxplot(potato$flavour~ potato$cook, data = potato, xlab = "Cooking method", ylab = "Flavor score")
# boxplot show similar variability----satisfy the assumption of homoscedasticity
# boxplot, each box can be considered as symmetric ---- satisfy the assumption of normality
# 160 groups of samples are independent ---- satisfy the assumption of independence

```

asdadsad



\section{Model Fitting}
\subsection{Definition of Terms}
In this section we define the terms used in the model fitting as shown in later stages.

\begin{table}[ht]
\centering
\label{table:terms}
\begin{tabular}{lc}
\toprule
Term & Explanation \\
\midrule
area\_grow & Growing area of potato \\
temp & Two week holding temperature \\
size\_potato & Size of potato \\
storage\_period & Storage period of potato \\
cooking\_method & Cooking method of potato \\
flavour & Flavour score of the cooked potato \\
\bottomrule
\end{tabular}
\caption{Term explanation of the Model}
\end{table}



\subsection{Least Squares and the Full Linear Model}
Least Squares (LS) is a parameter estimation method in regression analysis that minimizes the sum of the squared residuals. The LS method finds the coefficients that minimize the sum of the squared residuals, which is the difference between the observed values and the predicted values.

In this study, we aim to investigate the relationship between the flavor score and the growing area, two-week holding temperature, size of potato, storage period, and cooking method. The full linear model is as follows:
\begin {dmath}
\text{flavour} = \beta_0 + \beta_1\text{area\_grow} + \beta_2\text{temp} + \beta_3\text{size\_potato} + \beta_4\text{storage\_period} + \beta_5\text{cooking\_method} + \beta_{6}\text{area\_grow} \times \text{temp} + \beta_{7}\text{area\_grow} \times \text{size\_potato} + \beta_{8}\text{area\_grow} \times \text{storage\_period} + \beta_{9}\text{area\_grow} \times \text{cooking\_method} + \beta_{10}\text{temp} \times \text{size\_potato} + \beta_{11}\text{temp} \times \text{storage\_period} + \beta_{12}\text{temp} \times \text{cooking\_method} + \beta_{13}\text{size\_potato} \times \text{storage\_period} + \beta_{14}\text{size\_potato} \times \text{cooking\_method} + \beta_{15}\text{storage\_period} \times \text{cooking\_method} + \epsilon
\end{dmath}

where $\beta_0$ is the intercept, $\beta_1$ to $\beta_{15}$ are the coefficients of the main effects and interactions, and $\epsilon$ is the error term. The full model is fitted using the least squares method to estimate the coefficients of the model.

\subsection{Selecting the Effective Linear Model}
We apply a full ANOVA model with all the interactions and main effects to find out what are the effective terms. By fitting the full model, we obtain the following ANOVA table:


\begin{table}[ht]
\centering

\label{tab:anova_results}
\begin{tabular}{lcccccc}
\toprule
Source & Df & Sum Sq & Mean Sq & F value & Pr(>F) \\
\midrule
area_grow          & 1 & 0.529 & 0.5290 & 16.289 & 9.93e-05 *** \\
temp          & 1 & 1.089 & 1.0890 & 33.533 & 6.40e-08 *** \\
size_potato          & 1 & 0.000 & 0.0002 & 0.008  & 0.9302      \\
storage_period       & 3 & 2.024 & 0.6747 & 20.777 & 8.60e-11 *** \\
cooking_method          & 4 & 1.344 & 0.3359 & 10.342 & 3.57e-07 *** \\
area_grow:temp     & 1 & 0.020 & 0.0202 & 0.624  & 0.4314      \\
area_grow:size_potato     & 1 & 0.049 & 0.0490 & 1.509  & 0.2219      \\
area_grow:storage_period  & 3 & 0.287 & 0.0955 & 2.941  & 0.0362 *    \\
area_grow:cooking_method     & 4 & 0.166 & 0.0415 & 1.278  & 0.2829      \\
temp:size_potato     & 1 & 0.016 & 0.0160 & 0.493  & 0.4842      \\
temp:storage_period  & 3 & 0.937 & 0.3122 & 9.612  & 1.05e-05 *** \\
temp:cooking_method     & 4 & 0.123 & 0.0309 & 0.951  & 0.4376      \\
size_potato:storage_period  & 3 & 0.943 & 0.3144 & 9.682  & 9.72e-06 *** \\
size_potato:cooking_method     & 4 & 0.076 & 0.0190 & 0.585  & 0.6741      \\
storage_period:cooking_method  & 12& 1.447 & 0.1206 & 3.713  & 9.90e-05 *** \\
Residuals     & 113& 3.670& 0.0325 &        &             \\
\bottomrule
\end{tabular}
\caption{ANOVA Results on the Full Model}
\end{table}


By observing the ANOVA result and filtering out the insignificant ($p\_value >= 0.05$) terms, we obtain the effective linear model as follows:
\begin{dmath}
\text{flavour} =\beta_0 + \beta_1\text{area\_grow} + \beta_2\text{temp} + \beta_3\text{storage\_period} + \beta_4\text{cooking\_method} + \beta_5(\text{area\_grow} \times \text{storage\_period}) + \beta_6(\text{temp} \times \text{storage\_period}) + \beta_7(\text{size\_potato} \times \text{storage\_period}) + \beta_8(\text{storage\_period} \times \text{cooking\_method}) + \epsilon
\label{eq:effective_model}
\end{dmath}

\subsection{Fitting the Effective Model}
We apply the least squares method to fit the effective model and estimate the coefficients of the model. Below we show the estimated effective model.
\begin{dmath}
\hat{\text{flavour}} = 2.97 + 0.06 \times \text{area\_grow2} - 0.42 \times \text{temp2} - 0.058 \times \text{storage\_period2} - 0.20 \times \text{storage\_period3} 
 + 0.21 \times \text{storage\_period4} - 0.24 \times \text{cooking\_method2} - 0.013 \times \text{cooking\_method3} + 0 \times \text{cooking\_method4} 
 - 0.21 \times \text{cooking\_method5} - 0.05 \times \text{area\_grow2:storage\_period2} + 0.11 \times \text{area\_grow2:storage\_period3} + 0.17 \times \text{area\_grow2:storage\_period4} 
 + 0.29 \times \text{temp2:storage\_period2} + 0.33 \times \text{temp2:storage\_period3} + 0.41 \times \text{temp2:storage\_period4} 
 + 0.04 \times \text{storage\_period1:size\_potato2} - 0.01 \times \text{storage\_period2:size\_potato2} + 0.21 \times \text{storage\_period3:size\_potato2} 
 - 0.23 \times \text{storage\_period4:size\_potato2} + 0.31 \times \text{storage\_period2:cooking\_method2} + 0.20 \times \text{storage\_period3:cooking\_method2} 
 - 0.088 \times \text{storage\_period4:cooking\_method2} + 0.21 \times \text{storage\_period2:cooking\_method3} + 0.14 \times \text{storage\_period3:cooking\_method3} 
 - 0.18 \times \text{storage\_period4:cooking\_method3} + 0.18 \times \text{storage\_period2:cooking\_method4} + 0.14 \times \text{storage\_period3:cooking\_method4} 
 + 0.088 \times \text{storage\_period4:cooking\_method4} + 0.21 \times \text{storage\_period2:cooking\_method5} + 0.39 \times \text{storage\_period3:cooking\_method5} 
 - 0.28 \times \text{storage\_period4:cooking\_method5}
\label{eq:result_model}
\end{dmath}
In this equation, "variableLevel" represents the level of the variable, for example, area\_grow2 represents the second level of the area\_grow variable. The coefficients of the model are estimated using the least squares method.

```{r lm cropped, include=FALSE}
lm_cropped_result <- lm(flavour ~ area + temp + storage + cook + area:storage + temp:storage + size:storage + storage:cook, data=potato)
coef(lm_cropped_result)
```



```{r aov full, include=FALSE}
potato$area<-as.factor(potato$area)
potato$temp<-as.factor(potato$temp)
potato$size<-as.factor(potato$size)
potato$storage<-as.factor(potato$storage)
potato$cook<-as.factor(potato$cook)
# printing the full AOV model
result_full <- aov(flavour ~ (area + temp + size + storage + cook)^2, data=potato)
summary(result_full)
```

```{r aov effective, include=FALSE}
# We notice that there are something not significant, thus we remove them
result_cropped <- aov(flavour ~ area + temp + storage + cook + area:storage + temp:storage + size:storage + storage:cook, data=potato)
summary(result_cropped)
```

\section{Model Assessment}
\subsection {Model Assumptions}
The assumptions of the model are as follows:
\begin{enumerate}
    \item The errors have mean 0: $E(\epsilon_i) = 0$.
    
    \item Errors are homoscedastic (constant variance): $\text{Var}(\epsilon_i) = \sigma^2$, for all $i$.
    
    \item Errors are uncorrelated: $\text{Cov}(\epsilon_i, \epsilon_j) = 0$, for all $i \neq j$.
    
    \item Errors are normally distributed: $\epsilon_i \sim N(0, \sigma^2)$.
\end{enumerate}

\subsection{Graphical Assessment}
\label{assessment}
We assess the model assumptions by examining the residuals of the model. We plot the residuals against the predicted values to check for homoscedasticity and against the fitted values to check for normality.


```{r ANOVA_assessment, fig.width=10, fig.height=8}
layout(matrix(1:4,ncol=2))
plot(result_cropped)
```



\section{Conclusions}

In this study, we investigated the relationship between the flavor score of cooked potatoes and the growing area, two-week holding temperature, size of potato, storage period, and cooking method. We applied the ANOVA model to fit the full linear model, to find out the effective terms, and to fit the effective model with least square.

We find out that many of the interactions are not significant, and the effective model is presented in Equation \ref{eq:effective_model}, to obtain the relation between the flavour score and the effective terms, least square method is applied to estimate the coefficients of the model, resulting in Equatio \ref{eq:result_model}. Finally, we assess the model assumptions by examining the residuals of the model as seen in Section \ref{assessment}.

